---
title: "Assignment 2"
author: "Ryan Johnson"
date: "9/17/2020"
output: 
  html_document:
    toc: yes
    toc_depth: 3  
    toc_float: yes
    theme: cosmo
---
```{r load libraries, include=FALSE}
library(ggplot2)
library(sf)
library(rnaturalearth)
library(tidyverse)
library(ggspatial)
library(cartogram)
library(ggthemes)
```

# Assignment info DELETE ME
Select a large area from Table C-1 below, and create a map of the area, with its associated subareas,
using each of three different map projections. Indicate the map projection you’ve used for each. Identify a
variable and use at least three different methods to show how it varies among the subareas.
# DELETE THE ABOVE


### Description EDIT ME

*blah blah blah*

### Map of Europe - ____ projection

```{r example world map, message=FALSE}
world <- ne_countries(scale = 50, returnclass = "sf") 

world_map <- ggplot(world) +
  geom_sf() +
  theme_economist_white()

world_map
```

```{r europe reference, message=FALSE}
eur_countries <- world %>%
  filter(continent == "Europe")

eur_map <- ggplot(eur_countries) +
  geom_sf() +
  theme_economist_white()

eur_map
```

