---
title: 'VIS Assignment #4'
author: "Ryan Johnson"
date: "10/4/2020"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    theme: "paper"
---
```{r load libraries, include=FALSE}
library(osmdata)
library(opentripplanner)
library(tidyverse)
library(sf)
library(ggthemes)
library(ggspatial)
```

## DELETE ME PROMPT

Select between five and 30 locations within a city or metropolitan area. For each location, generate
isochrones for the same travel time for at least two different modes. Calculate the area of each isochrone
and compare the areas of the isochrones for the two (or more) modes you analyzed. Create three figures
(which may or may not be maps) to illustrate the results of your analysis.

## Description

I begin this exercise by loading a GeoJSON file from the City of Austin's Open Data Portal, which shows the locations of 20 community recreation centers throughout the city.

```{r read geojson, message=FALSE}
ATX_rec <- st_read("https://data.austintexas.gov/resource/8dff-2vkt.geojson")
```

Next I import the OpenStreetMap data for Austin's street network.

```{r sf street data, message=FALSE}
TxStateCent <- "+proj=lcc +lat_1=30.11666666666667 +lat_2=31.88333333333333 +lat_0=29.66666666666667 +lon_0=-100.3333333333333 +x_0=700000 +y_0=3000000 +ellps=GRS80 +datum=NAD83 +to_meter=0.3048006096012192 +no_defs"

austin_street_features <- opq(bbox = 'Austin TX USA') %>%
  add_osm_feature(key = 'highway') %>%
  osmdata_sf()

austin_streets <- austin_street_features$osm_lines %>%
  st_transform(crs = TxStateCent)
```

```{r street plot, message=FALSE, fig.width=8, fig.height=8}
ggplot(austin_streets) +
  geom_sf() +
  theme_map()
```

### Open Trip Planner setup

In order to create isochrones for various modes, I first installed the Open Trip Planner Java utility. 

```{r OTP graph, message=FALSE, results='hide'}
path_data <- file.path(getwd(), "OTP")

path_otp <- paste(path_data, "otp.jar", sep = "/")

otp_build_graph(otp = path_otp, dir = path_data, memory = 1024) 
```

After installing the utility, I'll finally launch OTP and connect to the open utility.

```{r OTP run, message=FALSE}
otp_setup(otp = path_otp, dir = path_data, memory =1024)

```

```{r OTP connect, message=FALSE}
otpcon <- otp_connect()

```

### Isochrones

```{r 5min isochrones}
iso_5min_walk <- 
  otp_isochrone(otpcon = otpcon, fromPlace = ATX_rec, 
                mode = "WALK", cutoffSec = 300) %>%
  st_transform(crs = TxStateCent) %>%
  mutate(mode = "walk")

iso_5min_drive <- 
  otp_isochrone(otpcon = otpcon, fromPlace = ATX_rec, 
                mode = "CAR", cutoffSec = 300) %>%
  st_transform(crs = TxStateCent) %>%
  mutate(mode = "drive")

iso_all_modes <- rbind(iso_5min_drive, iso_5min_walk)

otp_stop()
```

```{r map of isochrones, message=FALSE, fig.height=8, fig.width=8}
right_side <- st_bbox(iso_all_modes)$xmax
left_side  <- st_bbox(iso_all_modes)$xmin
top_side <- st_bbox(iso_all_modes)$ymax
bottom_side <- st_bbox(iso_all_modes)$ymin

ggplot(iso_all_modes) +
  annotation_map_tile(zoomin = 0, type = "cartolight", progress = "none") +
  geom_sf(aes(fill = mode), alpha = 0.5) +
  geom_sf(data = ATX_rec) +
  coord_sf(xlim = c(left_side, right_side), 
           ylim = c(bottom_side, top_side), expand = FALSE) +
  scale_fill_viridis_d(name = "Area reachable within 5 minutes",
                       labels = c("By car", "By foot")) +
  theme_map() +
  labs(caption = "Basemap Copyright OpenStreetMap contributors")
```






