---
title: "Assignment 3 VIS"
author: "Ryan Johnson"
date: "9/27/2020"
output: html_document
---

```{r libraries, message=FALSE, include=FALSE}
library(ggplot2)
library(tidyverse)
library(ggthemes)
library(ggspatial)
library(sf)
library(nngeo)
library(units)
```

Identify a municipal open data portal and find two point layers (A and B) and two polygon layers (C and
D).
Calculate six of the following 22 possible metrics, and illustrate the results of each calculation with a map.
Each of the four layers you selected should appear in at least one of those six maps.

## Description

Sources: https://data.austintexas.gov/ & https://data.texas.gov/

__Layers:__

A. CoA affordable housing inventory (point)
```{r loading housing inventory, message=FALSE, echo=FALSE}
Afford <- st_read("DataLoads927/AffordableIndexLayer.shp", quiet = TRUE)
```

B. Capital Metro transit stops (point) - includes only bus/train stops within Austin proper
```{r loading metro stops, message=FALSE, echo=FALSE}
CMTA <- st_read("DataLoads927/Stops.shp", quiet = TRUE) %>% 
  filter(CITY == "Austin")
```

C. CoA zoning map (polygon) - filtered to include only those parcels zoned for Planned Unit Development (PUD)
```{r loading and cleaning zoning, message=FALSE, echo=FALSE}
Zoning <- st_read("DataLoads927/geo_export_03a6167a-6c0b-498f-9e12-55f03963c9af.shp", quiet = TRUE)
```

```{r PUD cleaning, message=FALSE, echo=FALSE}
PUDs <- Zoning %>%
  filter(zoning_zty == "PUD")
```

D. CoA council districts (polygon)
```{r loading and council districts, message=FALSE, echo=FALSE}
Council <- st_read("DataLoads927/geo_export_a9775d2b-b31e-4b39-aa22-162244365c95.shp", quiet = TRUE) 
```

__Metrics:__

1. Average distance b/w A points and their nearest respective B points,
2. Number and proportion of A points in each D polygon,
3. Number and proportion of B points in each D polygon,
4. The number and proportion of C polygons containing A points,
5. The average density of B points in each D polygon, and
6. The average density of A points in each D polygon

```{r crs st plane central, include=FALSE}
TxStateCent <- "+proj=lcc +lat_1=30.11666666666667 +lat_2=31.88333333333333 +lat_0=29.66666666666667 +lon_0=-100.3333333333333 +x_0=700000 +y_0=3000000 +ellps=GRS80 +units=m +no_defs"
```

```{r transforming coordinate system, message=FALSE}

CMTA2 <- CMTA %>%
  st_transform(TxStateCent)

PUDs2 <- PUDs %>%
  st_transform(TxStateCent)

Council2 <- Council %>%
  st_transform(TxStateCent)

Afford2 <- Afford %>%
  st_transform(TxStateCent)
```

```{r example map, message=FALSE}
ggplot(Council2) +
  geom_sf(fill = "cornsilk", color = NA) +
  geom_sf(data = PUDs2, color = "lightblue", fill = "lightblue3") +
  geom_sf(data = Afford2, color = "magenta", size = 0.01) +
  geom_sf(data = CMTA2, color = "orange", size = 0.02) +
  theme_map() +
  annotation_scale()
```


## Metrics






