---
title: "Assignment 1 VIS"
author: "Ryan Johnson"
date: "9/13/2020"
output: html_document
---

```{r libraries, message=FALSE, results='hide', include=FALSE}
library(ggplot2)
library(tidyverse)
library(tidycensus)
library(viridis)
library(wesanderson)
library(ggsci)
```

```{r load household vars, message=FALSE, results='hide'}
hh_vars_2018 <- pums_variables %>%
  distinct(year, survey, var_code, 
           var_label, data_type, level) %>%
  filter(level == "housing", year == 2018, survey == "acs1")
```

##Defining the data set

Below, I define a data set including the following variables:

* **HINCP** - Household income - continuous
* **RNTP** - Monthly rent ($) - continuous
* **NP** - Number of people per household - continuous
* **BDSP** - Number of bedrooms per household - continuous
* **HHT_label** - Household type - categorical
* **BLDG_AGE_dec** - Age of building in decades - categorical

This sample began with all Texas responses from the 2018 ACS 1-year survey, and was further narrowed to include only those households with _some_ household income and monthly rent payment. Then, I added a filter to narrow my data to only those responses collected in Travis County. 

```{r PUMA-level data from Travis Co, TX, message=FALSE, results='hide'}
Travis_puma_rentals <- get_pums(variables = c(
                                      "PUMA", 
                                      "HINCP", 
                                      "RNTP", 
                                      "NP", 
                                      "BDSP",
                                      "HHT",
                                      "YBL"),
                        state = "TX",
                        year = 2018, 
                        survey = "acs1",
                        recode = TRUE) %>%
     mutate(PUMA = as.numeric(PUMA)) %>%
     mutate(BLDG_AGE_dec = case_when(
      YBL_label == "1939 or earlier" ~ 8,
      YBL_label == "1940 to 1949" ~ 7,
      YBL_label == "1950 to 1959" ~ 6,
      YBL_label == "1960 to 1969" ~ 5,
      YBL_label == "1970 to 1979" ~ 4,
      YBL_label == "1980 to 1989" ~ 3,
      YBL_label == "1990 to 1999" ~ 2,
      YBL_label == "2000 to 2004" ~ 1,
      YBL_label == "2005" ~ 1,
      YBL_label == "2006" ~ 1,
      YBL_label == "2007" ~ 1,
      YBL_label == "2008" ~ 1,
      YBL_label == "2009" ~ 1,
      YBL_label == "2010" ~ 0,
      YBL_label == "2011" ~ 0,
      YBL_label == "2012" ~ 0,
      YBL_label == "2013" ~ 0,
      YBL_label == "2014" ~ 0,
      YBL_label == "2015" ~ 0,
      YBL_label == "2016" ~ 0,
      YBL_label == "2017" ~ 0,
      YBL_label == "2018" ~ 0,)) %>% 
     filter(HINCP > 0, RNTP > 0, PUMA > 05300 & PUMA < 05310) %>%
     select(HINCP, RNTP, NP, BDSP, HHT_label, BLDG_AGE_dec)

```

Defining a subset of the above data, representing only the Central Austin PUMA:

```{r PUMA-level data from Central Austin only, message=FALSE, results='hide'}
Central_austin_rentals <- get_pums(variables = c(
                                      "PUMA", 
                                      "HINCP", 
                                      "RNTP", 
                                      "NP", 
                                      "BDSP",
                                      "HHT",
                                      "YBL"),
                        state = "TX",
                        year = 2018, 
                        survey = "acs1",
                        recode = TRUE) %>%
     mutate(PUMA = as.numeric(PUMA)) %>%
     mutate(BLDG_AGE_dec = case_when(
      YBL_label == "1939 or earlier" ~ 8,
      YBL_label == "1940 to 1949" ~ 7,
      YBL_label == "1950 to 1959" ~ 6,
      YBL_label == "1960 to 1969" ~ 5,
      YBL_label == "1970 to 1979" ~ 4,
      YBL_label == "1980 to 1989" ~ 3,
      YBL_label == "1990 to 1999" ~ 2,
      YBL_label == "2000 to 2004" ~ 1,
      YBL_label == "2005" ~ 1,
      YBL_label == "2006" ~ 1,
      YBL_label == "2007" ~ 1,
      YBL_label == "2008" ~ 1,
      YBL_label == "2009" ~ 1,
      YBL_label == "2010" ~ 0,
      YBL_label == "2011" ~ 0,
      YBL_label == "2012" ~ 0,
      YBL_label == "2013" ~ 0,
      YBL_label == "2014" ~ 0,
      YBL_label == "2015" ~ 0,
      YBL_label == "2016" ~ 0,
      YBL_label == "2017" ~ 0,
      YBL_label == "2018" ~ 0,)) %>% 
     filter(HINCP > 0, RNTP > 0, PUMA == 05306) %>%
     select(HINCP, RNTP, NP, BDSP, HHT_label, BLDG_AGE_dec)

```

Sample scatterplot:

```{r sample plot 1, message=FALSE, results='hide'}
ggplot(Travis_puma_rentals, 
       aes(x = RNTP,
           y = HINCP,
           color = BDSP)) +
  geom_point(size = 2) +
  stat_smooth(method = "lm") +
  scale_size_continuous(name = "Austin rent and income") +
  scale_x_continuous(name = "Monthly rent payment ($)",
                     breaks = seq(0, 10000, by = 500),
                     labels = paste(seq(0, 10000, by = 500))) +
  scale_y_continuous(name = "Household income within county ($)",
                     breaks = seq(0, 1000000, by = 50000)) +
  scale_color_continuous(name = "Bedrooms per unit") +
  theme_bw() +
  theme(legend.key.size = unit(0.5, "cm"))
```
```{r}
ggplot(Travis_puma_rentals, 
       aes(x = HINCP,
           y = RNTP)) +
  geom_point() +
  stat_smooth(color = "black", linetype = 2, size = 0.5) +
  scale_x_continuous(name = "Household income") +
  scale_y_continuous(name = "Monthly rent ($)") +
  theme_bw() 
```

```{r}
ggplot(Travis_puma_rentals, 
       aes(x = HINCP,
           y = RNTP)) +
  geom_point() +
  stat_smooth(color = "black", linetype = 2, size = 0.5) +
  scale_x_continuous(name = "Household income",
                             trans = "log",
                             breaks = c(1000, 10000, 100000, 1000000),
                             labels = c("1,000", "10,000", "100,000", "1,000,000")) +
  scale_y_continuous(name = "Monthly rent ($)") +
  theme_bw() 
```

```{r same as above but polar, message=FALSE, results='hide'}
ggplot(Travis_puma_rentals, 
       aes(x = RNTP,
           y = HINCP)) +
  geom_point() +
  stat_smooth(color = "black", linetype = 2, size = 0.5) +
  scale_y_continuous(name = "Household income",
                     trans = "log",
                     breaks = c(1000, 10000, 100000, 1000000),
                     labels = c("1,000", "10,000", "100,000", "1,000,000")) +
  scale_x_continuous(name = "Monthly rent ($)") +
  coord_polar(theta = "x")
  theme_bw() 
```

```{r}
ggplot(Travis_puma_rentals,
       aes(x = HHT_label,
           y = RNTP,
           color = HHT_label)) +
  geom_point(position = "jitter",
             size = 0.5,
             alpha = 0.75) +
  scale_x_discrete(name = "Household type", 
                   theme(axis.text.x = element_blank())) +
  scale_y_continuous(name = "Monthly rent ($)",
                     breaks = seq(0, 3000, by = 250)) +
  theme_bw()
```
```{r}
ggplot(Travis_puma_rentals,
       aes(x = HHT_label,
           y = RNTP,
           color = HHT_label)) +
  geom_boxplot() +
  scale_x_discrete(name = "Household type", 
                   theme(axis.text.x = element_blank())) +
  scale_y_continuous(name = "Monthly rent ($)",
                     breaks = seq(0, 3000, by = 250)) +
  theme_bw()
```

```{r}
pie_chart_1 <- ggplot(Travis_puma_rentals, 
                      aes(x = 1, fill = HHT_label)) +
  geom_bar(position = "fill") +
  coord_polar(theta = "y") +
  scale_y_continuous(name = "Number of households") +
  scale_fill_discrete(name = "Household type") +
  theme_void()

pie_chart_1
```

```{r, eval=FALSE}
ggplot(Travis_puma_rentals, 
       aes(x = NP,
           y = RNTP)) +
  geom_histogram(inherit.aes = TRUE) +
  scale_size_continuous(name = "Austin rent and household size") +
  scale_x_binned(name = "Number of people per household") +
  scale_y_continuous(name = "Monthly rent ($)") +
  theme_bw()
```

